Illustrating Model Properties

### 2024-05-28 manuscript.qmd
improve burnin display

## Requirements for a Bilateral Model

@@

We compared three different aprroaches to modelling the contralateral involvement in OPSCC:

1. A naive model of the contralateral side only. Neither the patterns of ipsilateral involvement, nor the information on whether the patients' tumors crossed the mid-sagittal line enter this model.
2. The bilateral model as introduced in the sections up to and including @sec-params-symmetry. The involvement patterns of both sides enter this model, but not whether the tumors extended the midline.
3. An additional bilateral model that tracks the midline extension of the patients' tumors as a random variable. This includes the formalism introduced up to and including @sec-midline. In contrast to the other two models, it does consider the reported midline extension in the data to attempt a more accurate prediction of the risk for contralateral occult disease.

@@

![Monitoring quantities during the burn-in phase of the parameter sampling.](figures/fig-model-burnin-history.png){#fig-model-burnin-history}

@fig-model-burnin-history is only a temporary placeholder for now.

@@

<!-- NOTE: This should probably belong to the Methods/Formalism section(s) -->

@@

### Midline Extension

@@

## Parameter Estimates

The predicted prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For a late T-category tumor, it is <!-- insert computational code later -->.

In the data, the prevalence of midline extension for early T-category tumors is <!-- insert computational code later -->. For late T-category tumors it is <!-- insert computational code later -->.

- Discrepancy between data and model is due to late T-stage's time prior
- still tied to other parameters
- cannot fully represent spread of midline extension and also represent spread params

>>>>

## Requirements for a Bilateral Model {#sec-requirements}

@@

In this section, we

@@

First, in @fig-model-burnin-history, we verify the sampling converged successfully by inspecting two monitoring quantities: The autocorrelation time of the MCMC chain and the acceptance fractions of the parallel walkers.

![Monitoring quantities during the burn-in phase of the parameter sampling. Left: The autocorrelation time of the sampling chain estimated at different sampling steps. We consider the chain converged when the estimate of the autocorrelation time is stable and drops below the trust threshold of $n/50$ where $n$ is the number of steps. Right: Fraction of accepted MCMC proposals averaged over all parallel walkers. Values around 30% indicate good mixing of the walkers.](figures/fig-model-burnin-history.png){#fig-model-burnin-history}

In @tbl-midline-params, we tabulate the mean and standard deviation of the sampled parameters for the full midline model.

@@

We want to investigate whether and to what extent the model can fulfill the requirements laid out in @sec-requirements. To that end, we compare its predictions for contralateral involvement against observations in the data. This is done given scenarios that differ in T-category and/or midline extension and/or ipsilateral involvement.

### Dependence of Contralateral Involvement on T-Category and Midline Extension

@@

### Prevalence of Midline Extension

@@

### 2024-05-28 model.qmd
improve burnin display

```{python}
#| echo: false
#| label: fig-model-burnin-history 
#| fig-cap: "Monitoring quantities during the burn-in phase of the parameter sampling."
from collections import namedtuple
import pandas as pd
from scripts import shared, paths

@@

nrows, ncols = 1, 3
label_map = {
  "acor_times": "autocorrelation [steps]",
  "accept_fracs": "acceptance fraction [%]",
  "max_log_probs": r"max. log-likelihood [$\times 10^3$]",
}

@@

for model in ["simple"]:
  history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
  for i, column in enumerate(history.columns):
    history.plot(y=column, ax=axes[i], label=model)
    axes[i].autoscale(enable=True, tight=True, axis="x")
    axes[i].set_ylabel(label_map[column])
    axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
    axes[i].set_xlabel(r"steps [$\times 10^3$]")

axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[2].yaxis.set_major_formatter(ticker.FuncFormatter(custom))

>>>>

```{python}
#| echo: false
#| label: fig-model-burnin-history 
#| fig-cap: "Monitoring quantities during the burn-in phase of the parameter sampling."
from collections import namedtuple
import pandas as pd
from scripts import shared, paths
import matplotlib.pyplot as plt
from matplotlib import ticker

params = shared.get_params()

@@

PlotConfig = namedtuple("PlotConfig", ["title", "ylabel", "label"])

nrows, ncols = 1, 2
config_map = {
  "acor_times": PlotConfig(
    title="Estimated Autocorrelation Time",
    ylabel="steps",
    label="Autocorrelation Time",
  ),
  "accept_fracs": PlotConfig(
    title="Average Acceptance Fraction of Walkers",
    ylabel=None,
    label=None,
  ),
}

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  width=17,
))

@@

for model in ["full"]:
  history = pd.read_csv(f"models/{model}/history.csv").set_index("steps")
  for i, (column, config) in enumerate(config_map.items()):
    axes[i].set_title(config.title, fontweight="bold")
    axes[i].plot(history.index, history[column], label=config.label)
    axes[i].autoscale(enable=True, tight=True, axis="x")
    axes[i].set_ylabel(config.ylabel)
    axes[i].xaxis.set_major_formatter(ticker.FuncFormatter(custom))
    axes[i].set_xlabel(r"steps [$\times 10^3$]")

axes[0].plot(
  history.index,
  history.index / params["sampling"]["trust_fac"],
  "--",
  color="black",
  label="Trust Threshold",
)
axes[0].legend()
axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
